---
title: "TidyTuesday Year 2025, Week 39"
format: html
---

These are my rough notes exploring this [weeks](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-09-30/readme.md) Tidy Tuesday dataset on crane observations at Lake Hornborgasjön in Sweden.


# Loading Library and Data

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(brms)
library(tidybayes)

cranes <- readr::read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-09-30/cranes.csv'
)
```


# EDA

- there seems to only be certain times of the year where the cranes are counted

plotting which parts of the year there are observations
```{r}
cranes |>
  mutate(
    year = year(date),
    day_of_year = yday(date)
  ) |>
  ggplot(aes(x = day_of_year, y = year, z = !is.na(date))) +
  geom_tile() +
  theme_classic()
```

- there is a seasonal pattern to the observed crane counts, with all the counts being made between ~day 75 - 120, and the second period being between ~day 225 - 300 
- the observations in the beginning of the year are more contiguous, vs the ones later in the year where observations are made every 2-3 days





plotting time series of crane observations, grouped by year
```{r}
cranes |>
  mutate(
    year = year(date),
    day_of_year = yday(date)
  ) |>
  ggplot(aes(x = day_of_year, y = observations, colour = year)) +
  geom_point() +
  theme_classic()
```

plotting faceted by year
```{r}
cranes |>
  mutate(
    year = year(date),
    day_of_year = yday(date)
  ) |>
  ggplot(aes(x = day_of_year, y = observations)) +
  geom_point() +
  facet_wrap(~year) +
  theme_classic()
```




# Modelling data

- Rather then treating the entire time series as a single series, each years worth of data will be stacked on top of each other, like the plots shown above and our "time" variable will just be the day of year.
- Each year have a pretty consistent pattern of crane observations, with two distinct peaks in the counts and I think capturing this will be easier with this approach, rather then trying to fit a single time series model to the entire dataset.

## Model 1
- Starting with a simple negative binomial regression model, with only day of year as a predictor
- Let $Y_{t,y}$ be the number of crane observations on day $t$ of year $y$
$$
Y_{t, y} \sim \text{NegBinom}(\lambda) \ \ \ \log(\lambda) = \beta_0 + \beta_1 t
$$



```{r}
model1_data <- cranes |>
  mutate(day_of_year = yday(date)) |>
  filter(!is.na(observations)) |>
  mutate(observations = observations)

model1 <- brm(
  data = model1_data,
  formula = observations ~ day_of_year,
  family = negbinomial(),
  cores = 4,
  seed = 123,
  backend = "cmdstan",
  refresh = 0
)
```


```{r}
model1 |> summary()
```

The R-hats, Bulk and ESS all look good


```{r}
pp_check(model1, ndraws = 100)
```

The posterior predictive check show the model follows the shape of the data, but doesn't capture the low counts well

```{r}
moddel1_loo <- loo(model1)
moddel1_loo
plot(moddel1_loo)
```

loo-cv check looks good, no problematic pareto k values


```{r}

tibble(day_of_year = seq(1, 365, by = 1)) |>
  add_epred_draws(model1, ndraws = 100) |>
  ggplot(aes(x = day_of_year)) +
  stat_lineribbon(aes(y = .epred), .width = c(.99, .95, .8, .5), alpha = 0.25) +
  geom_point(
    data = model1_data,
    aes(x = day_of_year, y = observations, colour = year(date)),
    inherit.aes = FALSE
  ) +
  scale_fill_brewer(palette = "Greys") +
  theme_classic()
``` 

This model doesn't come close to capturing the time dependent pattern of the data, but that was to be expected since this was a simple regression model. The next step is include the seasonality/time series nature of the data into the model. There are a couple of ways to do this:

* AR models
* splines
* gaussian processes
* finding the periodicities in the data and including them as predictors



## Model 2

- model 1, but with a spline on the day_of_year model

```{r}
model2_data <- cranes |>
  mutate(day_of_year = yday(date), year = year(date)) |>
  filter(!is.na(observations)) |>
  mutate(observations = observations)

model2 <- brm(
  data = model2_data,
  formula = observations ~ s(day_of_year),
  family = negbinomial(),
  cores = 4,
  seed = 123,
  backend = "cmdstan",
)
```


```{r}
model2 |> summary()

pp_check(model2, ndraws = 100)

loo(model1, model2)
```


```{r}
expand_grid(
  day_of_year = seq(1, 365, by = 1),
) |>
  add_epred_draws(model2, ndraws = 50) |>
  ggplot(aes(x = day_of_year)) +
  geom_point(
    data = model2_data,
    aes(x = day_of_year, y = observations, colour = year(date)),
    inherit.aes = FALSE
  ) +
  stat_lineribbon(aes(y = .epred), .width = c(.99, .95, .8, .5), alpha = 0.25) +
  scale_fill_brewer(palette = "Greys") +
  theme_classic()


expand_grid(
  day_of_year = seq(1, 365, by = 1),
  year = year(model2_data$date) |> unique()
) |>
  add_epred_draws(model2, ndraws = 50) |>
  ggplot(aes(x = day_of_year)) +
  geom_point(
    data = model2_data,
    aes(x = day_of_year, y = observations),
    inherit.aes = FALSE
  ) +
  stat_lineribbon(aes(y = .epred), .width = c(.99, .95, .8, .5), alpha = 0.25) +
  facet_wrap(~year) +
  scale_fill_brewer(palette = "Greys") +
  theme_classic()
``` 

plotting the predictions from the model on top of the true data, we see adding the splines definitely captured the rise and fall of the counts in the two time periods of the year where cranes were observed. However it doesn't capture the variance of the data, plotting the predictions facet by year shows that. but that may be because we haven't included the year variable into the model.


## Model 3
- model 2 but with year as a random effect
- using a factor-smooth interaction spline (fs) to allow the smooth for day_of_year to vary by year

```{r}
model3_data <- cranes |>
  mutate(day_of_year = yday(date), year = year(date)) |>
  filter(!is.na(observations)) |>
  mutate(observations = observations)

model3 <- brm(
  data = model3_data |> mutate(year = as.character(year)),
  formula = observations ~ s(day_of_year, year, bs = "fs"),
  family = negbinomial(),
  cores = 4,
  seed = 123,
  backend = "cmdstan",
)
```



```{r}
model3 |> summary()

pp_check(model3, ndraws = 100)

loo(model3)
```



```{r}
expand_grid(
  day_of_year = seq(1, 365, by = 1),
  year = c(2020, 2021, 2022)
) |>
  add_epred_draws(model3, ndraws = 20) |>
  ggplot(aes(x = day_of_year)) +
  geom_point(
    data = model3_data |> filter(year %in% c(2020, 2021, 2022)),
    aes(x = day_of_year, y = observations, colour = year),
    inherit.aes = FALSE
  ) +
  geom_line(
    aes(y = .epred, group = interaction(.draw, year)),
    alpha = 0.4,
    colour = "grey",
  ) +
  ylim(0, 20000) +
  # stat_lineribbon(aes(y = .epred), alpha = 0.25) +
  scale_fill_brewer(palette = "Greys") +
  facet_wrap(~year) +
  theme_classic()
```