---
title: "TidyTuesday Year 2025, Week 39"
format: html
---


# Loading Library and Data

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(brms)
library(tidybayes)

cranes <- readr::read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-09-30/cranes.csv'
)
```


# EDA

- there seems to only be certain times of the year where the cranes are counted

plotting which parts of the year there are observations
```{r}
cranes |>
  mutate(
    year = year(date),
    day_of_year = yday(date)
  ) |>
  ggplot(aes(x = day_of_year, y = year, z = !is.na(date))) +
  geom_tile() +
  theme_classic()
```

- there is a seasonal patter to the observed crane counts, with all the counts being made between ~day 75 - 120, and the second period being between ~day 225 - 300 
- the observations in the beginning of the year are more contiguous, vs the ones later in the year where observations are made every 2-3 days





plotting time series of crane observations, grouped by year
```{r}
cranes |>
  mutate(
    year = year(date),
    day_of_year = yday(date)
  ) |>
  ggplot(aes(x = day_of_year, y = observations, colour = year)) +
  geom_point() +
  theme_classic()
```


# Modelling data

## Model 1
- Starting with a simple negative binomial regression model, with only day of year as a predictor
- Let $Y_{t,y}$ be the number of crane observations on day $t$ of year $y$
$$
Y_{t, y} \sim \text{NegBinom}(\lambda) \ \ \ \log(\lambda) = \beta_0 + \beta_1 t
$$



```{r}
model1_data <- cranes |>
  mutate(day_of_year = yday(date)) |>
  filter(!is.na(observations)) |>
  mutate(observations = observations)

model1 <- brm(
  data = model1_data,
  formula = observations ~ day_of_year,
  family = negbinomial(),
  cores = 4,
  seed = 123,
  backend = "cmdstan"
)
```


```{r}
model1 |> summary()
```

The R-hats, Bulk and ESS all look good


```{r}
pp_check(model1, ndraws = 100)
```

The posterior predictive check show the model follows the shape of the data, but doesn't capture the low counts well

```{r}
moddel1_loo <- loo(model1)
moddel1_loo
plot(moddel1_loo)
```

loo-cv check looks good, no problematic pareto k values


```{r}

tibble(day_of_year = seq(1, 365, by = 1)) |>
  add_epred_draws(model1, ndraws = 100) |>
  ggplot(aes(x = day_of_year)) +
  stat_lineribbon(aes(y = .epred), .width = c(.99, .95, .8, .5), alpha = 0.25) +
  geom_point(
    data = model1_data,
    aes(x = day_of_year, y = observations, colour = year(date)),
    inherit.aes = FALSE
  ) +
  scale_fill_brewer(palette = "Greys") +
  theme_classic()
``` 

This model doesn't come close to capturing the time dependent pattern of the data, but that was to be expected since this was a simple regression model. The next step is include the seasonality/time series nature of the data into the model. There are a couple of ways to do this
- AR models
- splines
- gaussian processes
- finding the periodicities in the data and including them as predictors


